### 2016-08-16 intro.tex
rewriting/fleshing out syntax section
notes

\section{Introduction}
\label{sec:intro}
\tsnote{Skeleton text from old intro.  (Full version of old intro now
  in an appendix.}
We revisit the formalization of Bloom filters~\cite{xxx} and closely
related data structures as primitives with explicit security goals.
Recently, Naor and Yogev~\cite{xxx} (hereafter NY) gave such a
formalization, providing syntax for a generalized abstraction of
Bloom filters, as well as casting the classical Bloom filter
soundness condition as a security experiment admiting adaptive
adversaries.  To briefly summarize, a Bloom filter is a data
structure for compactly representing a set~$\mathcal{S}$ in a way
that supports constant-time responses to set-membership queries. The
representation of~$\mathcal{S}$ is a bit-array~$M$ of length~$m$,
all initially set to~0.  To insert $x \in \mathcal{S}$ into~$M$, one
computes indices $h_1(x),h_2(x),\ldots,h_k(x)$ where each $h_i
\colon \mathcal{U} \to \{0,1,\ldots m-1\}$ maps from a
universe~$\mathcal{U}$ (which contains $\mathcal{S}$) to indices
of~$M$.  The bits at positions $h_1(x),h_2(x),\ldots,h_k(x)$ are set
to 1; if a bit is already 1, due to inserting some previous element
of~$\mathcal{S}$, then it is left alone.  After all elements
of~$\mathcal{S}$ have been inserted, the resulting bit-array~$M$ is
taken as the representation.  To test if $y \in \mathcal{S}$, one
checks if \[(M[h_1(y)]=1 \wedge M[h_2(y)]=1 \wedge \cdots \wedge
M[h_k(y)]=1),\] returning ``yes'' if so, and ``no'' otherwise. Bloom
filters are correct, in the sense that if $y \in \mathcal{S}$, the
set-membership test always returns ``yes''.  The classical soundness
condition asserts that for all $\mathcal{S}$ and any $y
\notin\mathcal{S}$, we have $\Prob{h_1,h_2,\ldots,h_k \getsr
\mathrm{Func}(\mathcal{U},[m])\colon M[h_1(y)]=1 \wedge \cdots
\wedge M[h_k(y)]=1} \leq \epsilon$ where $\epsilon$, the
false-positive error, is a parameter of the Bloom filter.  (The
values of~$\epsilon,m,k$, and $|\mathcal{S}|$ are related in a
well-known way.) NY recast this soundness condition as a game in
which an adversary~$A$ is given~$\mathcal{S}$ and an oracle for
set-membership queries, and it attempts to find one or more false
positives. The fact that~$A$ can \emph{adapt} its queries and its
output based on what it learns from the set-membership oracle (but,
as we will see, not upon the hash functions) is the key definitional
contribution of~NY.

>>>>

\section{Introduction}
\label{sec:intro}
\tsnote{Skeleton text from old intro.  (Full version of old intro now
  in an appendix.} \jnote{Too focused on Bloom filters for my
  taste.}
We revisit the formalization of Bloom filters~\cite{xxx} and closely
related data structures as primitives with explicit security goals.
Recently, Naor and Yogev~\cite{xxx} (hereafter NY) gave such a
formalization, providing syntax for a generalized abstraction of
Bloom filters, as well as casting the classical Bloom filter
soundness condition as a security experiment admitting adaptive
adversaries.  To briefly summarize, a Bloom filter is a data
structure for compactly representing a set~$\mathcal{S}$ in a way
that supports constant-time responses to set-membership queries. The
representation of~$\mathcal{S}$ is a bit-array~$M$ of length~$m$,
all initially set to~0.  To insert $x \in \mathcal{S}$ into~$M$, one
computes indices $h_1(x),h_2(x),\ldots,h_k(x)$ where each $h_i
\colon \mathcal{U} \to \{0,1,\ldots m-1\}$ maps from a
universe~$\mathcal{U}$ (which contains $\mathcal{S}$) to indices
of~$M$.  The bits at positions $h_1(x),h_2(x),\ldots,h_k(x)$ are set
to 1; if a bit is already 1, due to inserting some previous element
of~$\mathcal{S}$, then it is left alone.  After all elements
of~$\mathcal{S}$ have been inserted, the resulting bit-array~$M$ is
taken as the representation.  To test if $y \in \mathcal{S}$, one
checks if
\[
(M[h_1(y)]=1 \wedge M[h_2(y)]=1 \wedge \cdots \wedge M[h_k(y)]=1),
\]
returning ``yes'' if so, and ``no'' otherwise. Bloom
filters are correct, in the sense that if $y \in \mathcal{S}$, the
set-membership test always returns ``yes''.  The classical soundness
condition asserts that for all $\mathcal{S}$ and any $y
\notin\mathcal{S}$, we have $\Prob{h_1,h_2,\ldots,h_k \getsr
\mathrm{Func}(\mathcal{U},[m])\colon M[h_1(y)]=1 \wedge \cdots
\wedge M[h_k(y)]=1} \leq \epsilon$ where $\epsilon$, the
false-positive error, is a parameter of the Bloom filter.  (The
values of~$\epsilon,m,k$, and $|\mathcal{S}|$ are related in a
well-known way.) NY recast this soundness condition as a game in
which an adversary~$A$ is given~$\mathcal{S}$ and an oracle for
set-membership queries, and it attempts to find one or more false
positives. The fact that~$A$ can \emph{adapt} its queries and its
output based on what it learns from the set-membership oracle (but,
as we will see, not upon the hash functions) is the key definitional
contribution of~NY.

### 2016-08-16 syntax.tex
notes

\section{(Set-multiplicity) Data Structures}

\heading{Preliminaries. }
When~$\univ$ is a set and $n>0$ is an integer, we let $[\univ]^n$ denote the set of all size-$n$ subsets of~$\univ$. We write $x \getsr \univ$ to denote sampling an element from~$\univ$ and assigning this to~$x$, and we extend this notation to randomized algorithms.  When~$X$ is a string over some alphabet~$\Sigma$, we write $|X|$ for the bitlength of~$X$, relative to some fixed encoding.  When~$X$ is a multiset, we overload the notation so that~$|X|$ is the number of elements in~$X$ (with mutiplicity).

@@

\heading{Data Structures.}  We being by defining a general primitive that produces a representation of a given multiset, and provides a mechanism to answer a specified collection of questions about that multiset.  Informally, the representation should serve as a good approximation of the multiset, at least with respect to the allowed questions.
\begin{definition}[Data Structure] \rm
Fix a non-empty set~$\univ$. A \emph{static data structure} (over universe~$\univ$)
is a tuple $\setprim=(\Sigma,\Gamma,\calS,\calQ,\Rep, \Qry)$ where:
\begin{itemize}
\item $\Sigma$ is the \emph{representation alphabet}
\item $\Gamma$ is the set of possible \emph{query answers}
\item $\calS$ is a set  of multisets over universe~$\univ$
\item $\mathcal{Q}=\{q \colon \mathcal{S} \to \Gamma\}$ is a set of supported \emph{queries}
\item $\Rep \colon \mathcal{S} \to \Sigma^* \times \bits^* \times \bits^*$ is a randomized \emph{representation algorithm}, taking as input $S \in \mathcal{S}$, and outputting a \emph{representation} $M \in \Sigma^*$, \emph{private side-information} $\privaux \in \bits^*$, and \emph{public side-information} $\pubaux \in \bits^*$
\item $\Qry\colon \Sigma^* \times \bits^* \times \calQ \to \Gamma$ is a deterministic \emph{query-evaluation} algorithm, taking as input $M, \privaux, q$, and returns \emph{answer} $a \in \Gamma$.
\end{itemize}
\hfill\dqed
\end{definition}
\tsnote{I implemented this change.}\tsnote{I'm now leaning towards having $\Rep$ output~$M$ and explicit pubic/private auxillary data.  Pushing secrets into~$M$ makes it more difficult to capture (for example) the Neidermeier et al.\ attacks in our privacy notions; see Figure~\ref{fig:privacy}.  In those attacks, the adversary would get the bitarray (which one naturally thinks of as~$M$) but not the secret keys for the PRFs/hash functions.  You could say that the bitarray is put in $\aux$ and that~$M$ \emph{only} contains the secret keys, but this is unnatural.}
All of the data structures that we will consider have the property that~$|M|$ does not depend on the coins of~$\Rep$. Thus we assume that~$|M|$ depends only on~$|S|$.  In general this need not be true, but the assumption does not restrict our syntax from capturing natural constructions (e.g., those seen in practice), and it makes it easier to state our security notions.

\begin{definition}[Error probability and rate] \rm
Fix a data structure $\setprim=(\Sigma,\Gamma,\calS,\calQ,\Rep, \Qry)$.
Let $\mathrm{Err}(S,q,M,\privaux)$ be a predicate that is true iff $\Qry(M, \privaux, q) \neq q(S)$.  Then a data structure
has \emph{error-probability} $\epsilon$ if
$\max_{S,q}\Pr[(M,\pubaux,\privaux) \getsr \Rep(S) : \mathrm{Err}(S,q,M,\privaux)=1 ] \leq \epsilon$
where the maximum is over $S \in \mathcal{S}$ and $q \in \mathcal{Q}$.  
%
Relatedly, it has \emph{error-rate} $\tilde{\epsilon}$ if
$\max_{S}\Pr[(M,\pubaux,\privaux) \getsr \Rep(S);\, q \getsr \calQ : \mathrm{Err}(S,q,M,\privaux)=1] \leq \tilde{\epsilon}$.
\hfill\dqed
\end{definition}

\noindent
The definition can be lifted to the random-oracle model
by giving $\Rep, \Qry$ access to a random oracle. In that case,
the probabilities defining the error-probability and error-rate are also taken over choice of the random oracle.

We will sometimes say that a data structure is $(1-\epsilon)$-correct, meaning it has error-probability~$\epsilon$.  Note that $1-\epsilon$ is a worst-case lower bound on the correctness of the data structure.

@@

\def\bin{{\sf Bin}}
\heading{Trivial attacks, and error-probability vs.\ error-rate. } 
Let~$\Pi$ be a scheme that is $(1-\epsilon)$-correct, and consider an attack that 
asks random queries $q_1,q_2,\ldots,q_T$, i.e., uniform samples from~$\calQ$.
%The expected number of random queries needed to
%find a error is $1/\epsilon$ tries; thus, an attacker making $O(t/\epsilon)$ queries succeeds with constant probability (almost always).
%Actually, it seems worthwhile to formalize this. 
Let $\mathrm{Err}_i$ be a random variable indicating the event $\Qry(M,\privaux,q_i)\neq q_i(S)$.  
The $(1-\epsilon)$-correctness of~$\Pi$ guarantees that the probability that $\mathrm{Err}_i=1$ is at most~$\epsilon$, 
and it is tempting to model the~$\mathrm{Err}_i$ as i.i.d. Bernoulli trials with success probability~$\epsilon$.
However, a moment's reflection shows that this is not the case.  Concretely, consider the following scheme.  Fix $|S|=n$ for all $S \in \calS$, where $n \ll |\univ|$.  For the queries, let $\calQ = \{q_x\colon \calS \to \bits \,|\, \forall x \in \univ\}$ where $\forall S \in \calS$ the predicate $q_x(S)=1 \Leftrightarrow x \in S$.  Let $\Rep(S)$ pick a random element~$s \in S$, and return $M = S \setminus \{s\}$ and $\privaux = \emptystring$.  To respond to queries, let~$\Qry(M,\privaux,q_x)=1$ iff $x \in M$.  It is easy to see that this scheme is $1/n$-correct; for any fixed~$S$, there are exactly~$n$ queries~$q_x$ that have probability~$1/n$  (over the coins of~$\Rep$) of causing a correctness error, and the remaining $|\univ|-n$ queries have probability zero of causing an error.  Moreover, once the coins of~$\Rep$ are fixed, there is exactly \emph{one} query that will cause an error.  Thus for random queries from~$Q$, for all~$i\in[T]$ we have $\Prob{\mathrm{Err}_i=1}=1/|\univ| \ll 1/n = \epsilon$.

This example is contrived, but it highlights the important difference
between error-probability and error-rate.  In particular, the
error-probability~$\epsilon$ is measured only over the coins
of~$\Rep$, for a fixed~$S$ and query~$q$.
Once the coins of~$\Rep$ are fixed, it is not straightforward how to
connect~$\epsilon$ to events that depend on other sources of
randomness. \tsnote{Not sure if there's more to say in this section, or if what's here is already too much.}

\heading{Set-multiplicity Data Structures. }
Our particular focus will be on data structures that support multiplicity queries, of which set-membership queries are a special case.  
\begin{definition} A \emph{set-multipicity data structure} is a data structure with $D=\mathbb{N}$ and where $\calQ=\{q_x \,|\, x \in \univ\}$ where $q_x(S)$ is defined to be the multiplicity of~$x$ in the multiset~$S$.
\hfill\dqed
\end{definition}
%
%\todo{Readdress issue of representation size.  A general statement might be something like `` The \emph{size} of a set-representation is the maximum (bit)length of any~$M$ that can be output by $\Rep$.''  But all of the natural constructions will have representation whose size will not depend on the coins of $\Rep$, i.e. they depend on $|S|$, number of hash functions, desired error rate. }

@@

%\heading{Bloom Filters and Related Data Structures. }
It is not hard to see Bloom filters are one possible instantiation a set-multiplicity data structure.
Fix an $n \geq 0$, let $\mathcal{S}=[\univ]^n$, and let $\Sigma=\Gamma=\bits$.  
Define the queries so that $q_x(S)=1$ if $x \in S$, and 0 otherwise.  Define algorithm~$\Rep$, with functions $k=k(|S|), m=m(|S|)$ and a description of a hash function family~$\mathcal{H}=\{h \colon \univ \to \{1,2,\ldots,m\}\}$ hardwired into it, to sample hash functions $h_1, \ldots, h_k$ from~$\mathcal{H}$ and then compute an $m$-bit array in the standard way.  It sets $\privaux=\pubaux = (h_1, \ldots, h_k)$ and~$M$ to be the bit array, and returns these.  The algorithm~$\Qry$ is defined so that, on input $(M,\privaux,q_x)$, it returns the minimum of the values (either 0 or 1) held at $M[h_1(x)],\ldots,M[h_k(x)]$.  We note that, while our error-probability and error-rate do not distinguish among ``types'' of errors, they also do not preclude instantiations (like the classical Bloom filter) that admit only one type of error, e.g. false-positives.

>>>>

\section{(Multiset) Data Structures}

\heading{Preliminaries. } When~$\univ$ is a set and $n>0$ is an
integer, we let $[\univ]^n$ denote the set of all size-$n$ subsets
\jnote{Multisets?} of~$\univ$. We write $x \getsr \univ$ to denote
sampling an element from~$\univ$ and assigning this to~$x$, and we
extend this notation to randomized algorithms.  When~$X$ is a string
over some alphabet~$\Sigma$, we write $|X|$ for the bitlength
of~$X$, relative to some fixed encoding.  When~$X$ is a multiset, we
overload the notation so that~$|X|$ is the number of elements in~$X$
(with mutiplicity).

@@

\heading{Data structures.}  We being by defining a general primitive
that produces a representation of a given multiset, and provides a
mechanism to answer a specified collection of questions about that
multiset.  Informally, the representation should serve as a good
approximation of the multiset, at least with respect to the allowed
questions.
\begin{definition}[Data structure] \rm
Fix a non-empty set~$\univ$. A \emph{static data structure} (over universe~$\univ$)
is a tuple $\setprim=(\Sigma,\Gamma,\calS,\calQ,\Rep, \Qry)$ where:
\begin{itemize}
\item $\Sigma$ is the \emph{representation alphabet}. \jnote{Is
    there any reason not to simply fix this to $\{0,1\}$?}
\item $\Gamma$ is the set of possible \emph{query answers}.
    \jnote{Do we need this for any reason?}
\item $\calS$ is a set  of multisets over universe~$\univ$.
\item $\mathcal{Q}=\{q \colon \mathcal{S} \to \Gamma\}$ is a set
    of supported \emph{queries}.
\item $\Rep \colon \mathcal{S} \to \Sigma^* \times \bits^*
    \times \bits^*$ is a randomized \emph{representation
    algorithm}, taking as input $S \in \mathcal{S}$, and
    outputting a \emph{representation} $M \in \Sigma^*$,
    \emph{private side information} $\privaux \in \bits^*$, and
    \emph{public side information} $\pubaux \in \bits^*$.
\item $\Qry\colon \Sigma^* \times \bits^* \times \calQ \to
    \Gamma$ is a deterministic \emph{query-evaluation}
    algorithm, taking as input $M, \privaux, q$, and returns
    \emph{answer} $a \in \Gamma$. \jnote{Should take $\pubaux$
    as input also.}
\end{itemize}
\hfill\dqed
\end{definition}
\tsnote{I implemented this change.}\tsnote{I'm now leaning towards
having $\Rep$ output~$M$ and explicit pubic/private auxillary data.
Pushing secrets into~$M$ makes it more difficult to capture (for
example) the Neidermeier et al.\ attacks in our privacy notions; see
Figure~\ref{fig:privacy}.  In those attacks, the adversary would get
the bitarray (which one naturally thinks of as~$M$) but not the
secret keys for the PRFs/hash functions.  You could say that the
bitarray is put in $\aux$ and that~$M$ \emph{only} contains the
secret keys, but this is unnatural.} All of the data structures that
we will consider have the property that~$|M|$ does not depend on the
coins of~$\Rep$. Thus we assume that~$|M|$ depends only on~$|S|$. In
general this need not be true, but the assumption does not restrict
our syntax from capturing natural constructions (e.g., those used in
practice), and it makes it easier to state our security notions.

\begin{definition}[Error probability and rate] \rm
Fix a data structure $\setprim=(\Sigma,\Gamma,\calS,\calQ,\Rep,
\Qry)$. Let $\mathrm{Err}(S,q,M,\privaux)$ be a predicate that is
true iff $\Qry(M, \privaux, q) \neq q(S)$.  Then a data structure
has \emph{error probability} $\epsilon$ if
$\max_{S,q}\Pr[(M,\pubaux,\privaux) \getsr \Rep(S) :
\mathrm{Err}(S,q,M,\privaux)=1 ] \leq \epsilon$.
%
It has \emph{error rate} $\tilde{\epsilon}$ if
$\max_{S}\Pr[(M,\pubaux,\privaux) \getsr \Rep(S);\, q \getsr \calQ :
\mathrm{Err}(S,q,M,\privaux)=1] \leq \tilde{\epsilon}$. \hfill\dqed
\end{definition}

\noindent The definition can be lifted to the random-oracle model by
giving $\Rep, \Qry$ access to a random oracle. In that case, the
probabilities defining the error probability and error rate are also
taken over choice of the random oracle.

We will sometimes say that a data structure is
$(1-\epsilon)$-correct, meaning it has error probability~$\epsilon$.
Note that $1-\epsilon$ is a worst-case lower bound on the
correctness of the data structure.

@@

\def\bin{{\sf Bin}}
\heading{Trivial attacks, and error probability vs.\ error rate. }
Let~$\Pi$ be a scheme that is $(1-\epsilon)$-correct, and consider
an attack that asks random queries $q_1,q_2,\ldots,q_T$, i.e.,
uniform samples from~$\calQ$.
%The expected number of random queries needed to
%find a error is $1/\epsilon$ tries; thus, an attacker making $O(t/\epsilon)$ queries succeeds with constant probability (almost always).
%Actually, it seems worthwhile to formalize this.
Let $\mathrm{Err}_i$ be a random variable indicating the event $\Qry(M,\privaux,q_i)\neq q_i(S)$.
The $(1-\epsilon)$-correctness of~$\Pi$ guarantees that the probability that $\mathrm{Err}_i=1$ is at most~$\epsilon$,
and it is tempting to model the~$\mathrm{Err}_i$ as i.i.d. Bernoulli trials with success probability~$\epsilon$.
However, a moment's reflection shows that this is not the case.  Concretely, consider the following scheme.  Fix $|S|=n$ for all $S \in \calS$, where $n \ll |\univ|$.  For the queries, let $\calQ = \{q_x\colon \calS \to \bits \,|\, \forall x \in \univ\}$ where $\forall S \in \calS$ the predicate $q_x(S)=1 \Leftrightarrow x \in S$.  Let $\Rep(S)$ pick a random element~$s \in S$, and return $M = S \setminus \{s\}$ and $\privaux = \emptystring$.  To respond to queries, let~$\Qry(M,\privaux,q_x)=1$ iff $x \in M$.  It is easy to see that this scheme is $1/n$-correct; for any fixed~$S$, there are exactly~$n$ queries~$q_x$ that have probability~$1/n$  (over the coins of~$\Rep$) of causing a correctness error, and the remaining $|\univ|-n$ queries have probability zero of causing an error.  Moreover, once the coins of~$\Rep$ are fixed, there is exactly \emph{one} query that will cause an error.  Thus for random queries from~$Q$, for all~$i\in[T]$ we have $\Prob{\mathrm{Err}_i=1}=1/|\univ| \ll 1/n = \epsilon$.

This example is contrived, but it highlights the important
difference between error probability and error rate.  In particular,
the error probability~$\epsilon$ is measured only over the coins
of~$\Rep$, for a fixed~$S$ and query~$q$. Once the coins of~$\Rep$
are fixed, it is not straightforward how to connect~$\epsilon$ to
events that depend on other sources of randomness. \tsnote{Not sure
if there's more to say in this section, or if what's here is already
too much.}

\heading{Set-multiplicity Data Structures. } Our particular focus
will be on data structures that support multiplicity queries, of
which set-membership queries are a special case.
\begin{definition} A \emph{set-multiplicity data structure} is a data
structure with $D=\mathbb{N}$ \jnote{Notation. (Anyway, I don't
think we need an explicit symbol for the range of queries.)} and
where $\calQ=\{q_x \,|\, x \in \univ\}$ where $q_x(S)$ is defined to
be the multiplicity of~$x$ in the multiset~$S$. \hfill\dqed
\end{definition}
%
%\todo{Readdress issue of representation size.  A general statement might be something like `` The \emph{size} of a set-representation is the maximum (bit)length of any~$M$ that can be output by $\Rep$.''  But all of the natural constructions will have representation whose size will not depend on the coins of $\Rep$, i.e. they depend on $|S|$, number of hash functions, desired error rate. }

@@

%\heading{Bloom Filters and Related Data Structures. }
It is not hard to see Bloom filters are one possible instantiation a
set-multiplicity data structure. Fix an $n \geq 0$, let
$\mathcal{S}=[\univ]^n$, and let $\Sigma=\Gamma=\bits$. Define the
queries so that $q_x(S)=1$ if $x \in S$, and 0 otherwise.  Define
algorithm~$\Rep$, with functions $k=k(|S|), m=m(|S|)$ and a
description of a hash function family~$\mathcal{H}=\{h \colon \univ
\to \{1,2,\ldots,m\}\}$ hardwired into it, to sample hash functions
$h_1, \ldots, h_k$ from~$\mathcal{H}$ and then compute an $m$-bit
array in the standard way.  It sets $\privaux=\pubaux = (h_1,
\ldots, h_k)$ and~$M$ to be the bit array, and returns these.  The
algorithm~$\Qry$ is defined so that, on input $(M,\privaux,q_x)$, it
returns the AND of the values (either 0 or 1) held at
$M[h_1(x)],\ldots,M[h_k(x)]$.  We note that, while our error
probability and error rate do not distinguish among ``types'' of
errors, they also do not preclude instantiations (like the classical
Bloom filter) that admit only one type of error, e.g.
false-positives.
